---
title: "sds_git_basics"
author: "Dulcie Vousden, DataKind UK"
date: "23/03/2020"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=FALSE)
```

```{r, echo=FALSE}
setwd("~/Desktop/")
```

## First things first

Much of this tutorial was taken from the full git tutorial from software carpentry: https://swcarpentry.github.io/git-novice/ (under create commons license). It will be more detailed and more informative than this one!



## Getting started

When we use Git on a new computer for the first time,
we need to configure a few things. Below are a few examples
of configurations we will set as we get started with Git:

*   our name and email address,
*   what our preferred text editor is,
*   and that we want to use these settings globally (i.e. for every project).

On a command line, Git commands are written as `git verb options`,
where `verb` is what we actually want to do and `options` is additional optional information which may be needed for the `verb`. 

```{bash}
git config --global user.name "Dulcie Vousden"
git config --global user.email "dulcievousden@gmail.com"
```


Please use your own name and email address instead of mine. If you already have a github account, this should be the same email address you used for that.
If you don't ignore this.
This user name and email will be associated with your subsequent Git activity,
which means that any changes pushed to
[GitHub](https://github.com/),
[BitBucket](https://bitbucket.org/),
[GitLab](https://gitlab.com/) or
another Git host server
in a later lesson will include this information.



Once Git is configured,
we can start using it.

## Creating a repository

First, let's create a directory in `Desktop` folder for our work and then move into that directory:

```{bash}
cd ~/Desktop
mkdir to_do
cd to_do
pwd
```


Then we tell Git to make `to_do` a [repository]({{ page.root }}{% link reference.md %}#repository)—a place where Git can store versions of our files:
A Git repository is a place where git can store versions of your files.  It stores both the files and the changes to the files.
To set up a repository we use git init
```{bash}
git init
```


It is important to note that `git init` will create a repository that
includes subdirectories and their files---there is no need to create
separate repositories nested within the `to_do` repository, whether
subdirectories are present from the beginning or added later. Also, note
that the creation of the `to_do` directory and its initialization as a
repository are completely separate processes.

If we use `ls` to show the directory's contents,
it appears that nothing has changed:

```{bash}
ls
```


But if we add the `-a` flag to show everything,
we can see that Git has created a hidden directory within `to_do` called `.git`:

```{bash}
ls -a
```


Git uses this special subdirectory to store all the information about the project, 
including all files and sub-directories located within the project's directory.
If we ever delete the `.git` subdirectory,
we will lose the project's history.

We can check that everything is set up correctly
by asking Git to tell us the status of our project:

```{bash}
git status
```

If you are using a different version of `git`, the exact
wording of the output might be slightly different.

## Adding and tracking files

Let's create a file called `shopping.txt` that contains some notes
We'll use `nano` to edit the file;
you can use whatever editor you like.


```{bash}
touch shopping.txt
```



Type the text below into the `shopping.txt` file:

**loo roll**

`shopping.txt` contains a single line, which we can see by running:

```{bash}
cat shopping.txt
```



If we check the status of our project again,
Git tells us that it's noticed the new file:

```{bash}
git status
```



The "untracked files" message means that there's a file in the directory
that Git isn't keeping track of.
We can tell Git to track a file using `git add`:

```{bash}
git add shopping.txt
```


and then check that the right thing happened:

```{bash}
git status
```



Git now knows that it's supposed to keep track of `shopping.txt`,
but it hasn't recorded these changes as a commit yet.
To get it to do that,
we need to run one more command:

```{bash}
git commit -m "Started shopping list"
```


When we run `git commit`,
Git takes everything we have told it to save by using `git add`
and stores a copy permanently inside the special `.git` directory.
This permanent copy is called a  commit

We use the `-m` flag (for "message")
to record a short, descriptive, and specific comment that will help us remember later on what we did and why.
If we just run `git commit` without the `-m` option,
Git will launch `nano` (or whatever other editor we configured as `core.editor`)
so that we can write a longer message.

[Good commit messages][commit-messages] start with a brief (<50 characters) statement about the
changes made in the commit. Generally, the message should complete the sentence "If applied, this commit will" <commit message here>.
If you want to go into more detail, add a blank line between the summary line and your additional notes. Use this additional space to explain why you made changes and/or what their impact will be.

If we run `git status` now:

```{bash}
git status
```


it tells us everything is up to date.
If we want to know what we've done recently,
we can ask Git to show us the project's history using `git log`:

```{bash}
git log
```



`git log` lists all commits  made to a repository in reverse chronological order.
The listing for each commit includes
the commit's full identifier
(which starts with the same characters as
the short identifier printed by the `git commit` command earlier),
the commit's author,
when it was created,
and the log message Git was given when the commit was created.


Now suppose you add more information to the file. Using nano, add **soap** to the shopping list


```{bash}
git status
```




```{bash}
git commit -m "added item to shopping list"
git status
```



Whoops:
Git won't commit because we didn't use `git add` first.
Let's fix that:

```{bash}
git add shopping.txt
git commit -m "added item to shopping list"
```




Git insists that we add files to the set we want to commit
before actually committing anything. 
To allow for this,
Git has a special *staging area*
where it keeps track of things that have been added to
the current changeset but not commited yet





To recap, when we want to add changes to our repository,
we first need to add the changed files to the staging area
(`git add`) and then commit the staged changes to the
repository (`git commit`)


## Remote repositories


Version control really comes into its own when we begin to collaborate with
other people.  We already have most of the machinery we need to do this; the
only thing missing is to copy changes from one repository to another.

Systems like Git allow us to move work between any two repositories.  In
practice, though, it's easiest to use one copy as a central hub, and to keep it
on the web rather than on someone's laptop.  Most programmers use hosting
services like [GitHub](https://github.com), [Bitbucket](https://bitbucket.org) or
[GitLab](https://gitlab.com/) to hold those master copies

Let's start by sharing the changes we've made to our current project with the
world. 

Go to GitHub: www.github.com and create an account
You don't have to do this part if you don't want to

If you do, use the same email address as you set in the git config
- you can ignore the survey
- you'll have to confirm in your email

Log in to GitHub, then click on the icon in the top right corner to
create a new repository called `to_do`:

Name your repository "to_do" and then click "Create Repository".

Note: Since this repository will be connected to a local repository, it needs to be empty. Leave 
"Initialize this repository with a README" unchecked, and keep "None" as options for both "Add 
.gitignore" and "Add a license." See the "GitHub License and README files" exercise below for a full 
explanation of why the repository needs to be empty.


As soon as the repository is created, GitHub displays a page with a URL and some
information on how to configure your local repository:

This effectively does the following on GitHub's servers:

- mkdir to_do
- cd to_do
- git init


Note that our local repository still contains our earlier work on `shopping.txt`, but the
remote repository on GitHub appears empty as it doesn't contain any files yet.

The next step is to connect the two repositories.  We do this by making the
GitHub repository a remote  for the local repository.
The home page of the repository on GitHub includes the string we need to
identify it:


```{bash}
git remote add origin https://github.com/dvousden/to_do.git
```


Make sure to use the URL for your repository rather than mine: the only
difference should be your username instead of `dvousden`.

`origin` is a local name used to refer to the remote repository. It could be called
anything, but `origin` is a convention that is often used by default in git
and GitHub, so it's helpful to stick with this unless there's a reason not to.

We can check that the command has worked by running `git remote -v`:

```{bash}
git remote -v
```


Once the remote is set up, this command will push the changes from
our local repository to the repository on GitHub:

```{bash}
git push origin master
```




We can pull changes from the remote repository to the local one as well:

```{bash}
git pull origin master
```



Pulling has no effect in this case because the two repositories are already
synchronized.  If someone else had pushed some changes to the repository on
GitHub, though, this command would download them to our local repository.


Lastly, let's suppose you want to download someone else's code. You can do this using git clone

```{bash}
git clone https://github.com/dvousden/sds_git_tutorial.git ~/Desktop/sds_git_tutorial/
```

Then, depending on the owner's settings, you can make changes and push them to the owner's repository.
Or alternatively, you can make your own branch, make changes, and choose when to merge them later.

The trick is that with git clone you've copied all these notes. :-)

Back to presentation!


## Key points

- "`git init` initializes a repository."
- "Git stores all of its repository data in the `.git` directory."
- '`git status` shows the status of a repository.'
- Files can be stored in a project's working directory (which users see), the staging
  area (where the next commit is being built up) and the local repository (where commits
  are permanently recorded).
- '`git add` puts files in the staging area.'
- '`git commit` saves the staged content as a new commit in the local repository.'
- "A local Git repository can be connected to one or more remote repositories."
- "Use the HTTPS protocol to connect to remote repositories until you have learned how to set up SSH."
- "`git push` copies changes from a local repository to a remote repository."
- "`git pull` copies changes from a remote repository to a local repository."
- "`git clone` copies a remote repository to create a local repository with a remote called `origin` automatically set up."